MS-A1 Formulae and Equations

### 2021-01-21 ms-a1_formulae-and-equations.rmd
Added MS-A1_c and MS-A1_d

* In $\dfrac{2x}{3}=\dfrac{x}{4}$ the lowest common denominator is $\ \ 12$.
* In $\dfrac{3}{x}=\dfrac{5}{2x-1}$ the lowest common denominator is $\ \ x(2x-1)$.

@@

>>>>

* In $\ \ \dfrac{2x}{3}=\dfrac{x}{4}\ \ $ the (lowest) common denominator is $\ \ 12$.
* In $\ \ \dfrac{3}{x}=\dfrac{5}{2x-1}\ \ $ the common denominator is $\ \ x(2x-1)$.

@@

# Formulae [![](pres.svg){width=16}](https://claytonpollard.github.io/stage6_standard_presentations/MS-A1/c_formulae.pdf){target="blank"} {.tabset .tabset-fade .tabset-pills}

* review evaluating the subject of a formula, given the value of other pronumerals in the formula

### Formulae

A **formula** is an equation which connects two or more variables. The plural of formula is **formulae** or **formulas**.

Generally a formula is written with a single variable, the **subject** on the LHS and the rest on the right.

$$\underset{\overset{\uparrow}{\text{subject}}}{A}=\frac{1}{2}ab\cos{C}$$

### Formula substitution

If the formula contains two or more variables and we know the value of all but one of them, we can solve an equation to find the remaining variable.

1) State the variables in the formula and the known values.
2) Write down the formula and Substitute the known values into the formula.
3) Solve the one variable equation to find the unknown value.

The area of a trapezium is given by $\ \ A=\dfrac{h}{2}(a+b)\ \ $ where $a$ and $b$ are the lengths of the parallel sides and $h$ is the height.

Calculate the height of a 20 cm$^2$ trapezium with parallel lengths of 3 cm and 7 cm.

The area of a trapezium is given by $\ \ A=\dfrac{h}{2}(a+b)\ \ $ where $a$ and $b$ are the lengths of the parallel sides and $h$ is the height.

Calculate the height of a 20 cm$^2$ trapezium with parallel lengths of 3 cm and 7 cm.

<div class="row">
  <div class="column">
  $\begin{aligned}A&=20\text{ cm}^2\\h&=\ ?\\a&=3\text{ cm}^2\\b&=7\text{ cm}^2\end{aligned}$
  </div>
  <div class="column">
  $\begin{aligned}A&=\frac{h}{2}(a+b)\\20&=\frac{h}{2}(3+7)\\20&=\frac{h}{2}(10)\\20&=5h\\20\color{red}{\div5}&=5h\color{red}{\div5}\\4&=h\\h&=4\text{ cm}\end{aligned}$
  </div>
</div>

The height of the trapezium is equal to 4 cm.

## Hide
# Rearranging Formulae [![](pres.svg){width=16}](https://claytonpollard.github.io/stage6_standard_presentations/MS-A1/d_rearranging-formulae.pdf){target="blank"} {.tabset .tabset-fade .tabset-pills}

* change the subject of a formula

### Formula Rearrangement

By performing operations to both sides of a formula, they can be **rearranged** to make **equivalent** formulae where other variables are the subjects.

<div class="row">
  <div class="column">
  $$\begin{aligned}V&=\pi r^2h\\V\color{red}{\div\pi r^2}&=\pi r^2h\color{red}{\div\pi r^2}\\\frac{V}{\pi r^2}&=h\\h&=\frac{V}{\pi r^2}\end{aligned}$$
  </div>
  <div class="column">
  $$\begin{aligned}V&=\pi r^2h\\V\color{red}{\div\pi h}&=\pi r^2h\color{red}{\div\pi h}\\\frac{V}{\pi h}&=r^2\\r^2&=\frac{V}{\pi h}\\\color{red}{\sqrt{\color{black}{r^2}}}&=\color{red}{\sqrt{\color{black}{\frac{V}{\pi h}}}}\\r&=\sqrt{\frac{V}{\pi h}}&(r>0)\end{aligned}$$
  </div>
</div>

Make $a$ the subject of $\quad2a-7b=23$.

Make $a$ the subject of $\quad2a-7b=23$.

$\begin{aligned}2a-7b&=23\\2a-7b\color{red}{+7b}&=23\color{red}{+7b}\\2a&=23-7b\\\color{red}{\frac{\color{black}{2a}}{2}}&=\color{red}{\frac{\color{black}{23-7b}}{2}}\\a&=\frac{23-7b}{2}\end{aligned}$

### Rearranging and substituting

Previously, during formula substitution, the variables were replaced by numbers and then the equation was solved. However, often we need to substitute several values for the unknowns and solve the equation for each case. In this situation it is quicker to **rearrange** the formula **before substituting**.

The surface area of a sphere is given by $\ \ A=4\pi r^2\ \ $ where $r$ is the sphere's radius.

a) Rearrange this formula to make $r$ the subject.
b) Hence find the radius to 3 significant figures when the surface area is:
    i) 10 cm$^2$
    ii) 20 cm$^2$
    iii) 30 cm$^3$

The surface area of a sphere is given by $\ \ A=4\pi r^2\ \ $ where $r$ is the sphere's radius.

a) Rearrange this formula to make $r$ the subject.
b) Hence find the radius to 3 significant figures when the surface area is:
    i) 10 cm$^2$
    ii) 20 cm$^2$
    iii) 30 cm$^3$

$\begin{aligned}A&=4\pi r^2\\A\color{red}{\div4\pi}&=4\pi r^2\color{red}{\div4\pi}\\\frac{A}{4\pi}&=r^2\\\color{red}{\sqrt{\color{black}{\frac{A}{4\pi}}}}&=\color{red}{\sqrt{\color{black}{r^2}}}\\r&=\sqrt{\frac{A}{4\pi}}&(r>0)\end{aligned}$

The surface area of a sphere is given by $\ \ A=4\pi r^2\ \ $ where $r$ is the sphere's radius.

a) Rearrange this formula to make $r$ the subject.
b) Hence find the radius to 3 significant figures when the surface area is:
    i) 10 cm$^2$
    ii) 20 cm$^2$
    iii) 30 cm$^3$

$\begin{aligned}\text{i) }\ \ r&=\sqrt{\frac{A}{4\pi}}&\text{ii) }\ \ r&=\sqrt{\frac{A}{4\pi}}&\text{iii) }\ \ r&=\sqrt{\frac{A}{4\pi}}\\\\r&=\sqrt{\frac{\color{red}{10}}{4\pi}}&r&=\sqrt{\frac{\color{red}{20}}{4\pi}}&r&=\sqrt{\frac{\color{red}{30}}{4\pi}}\\\\r&=0.892\text{ cm}&r&=1.26\text{ cm}&r&=1.55\text{ cm}\end{aligned}$
